---
title: "Emotion Detection with graniteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Emotion Detection with graniteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 10,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates using graniteR for multi-class emotion detection in text data. We use the emotion dataset from HuggingFace (dair-ai/emotion), which contains text samples labeled with six basic emotions based on Ekman's emotion theory.

The dataset includes 20,000 labeled examples, making it suitable for training robust emotion classifiers. This use case illustrates graniteR's capabilities in:

- Multi-class text classification (6 classes)
- Working with imbalanced datasets
- Fine-tuning transformer models for emotion analysis
- Evaluating model performance across multiple classes

## Prerequisites

Before proceeding, ensure you have installed graniteR and its Python dependencies using UV:

```bash
# From the graniteR package directory
./inst/bash/setup_python.sh

# Then in R, configure the Python environment
Sys.setenv(RETICULATE_PYTHON = ".venv/bin/python")
```

UV provides fast dependency installation (1-2 minutes). See the "Getting Started" vignette for detailed installation instructions.

For this classification task, GPU acceleration is recommended but not required. If you have a CUDA-capable GPU, the training will be significantly faster.

## Understanding the Task

Emotion detection is the task of identifying the emotional tone expressed in text. Unlike simple binary sentiment analysis (positive/negative), emotion detection classifies text into specific emotional categories:

- **Sadness**: Expressing sorrow, disappointment, or grief
- **Joy**: Expressing happiness, pleasure, or satisfaction
- **Love**: Expressing affection, care, or romantic feelings
- **Anger**: Expressing frustration, rage, or displeasure
- **Fear**: Expressing worry, anxiety, or concern
- **Surprise**: Expressing unexpectedness or astonishment

This fine-grained emotion classification is useful for:
- Customer feedback analysis
- Mental health monitoring
- Social media sentiment tracking
- Chatbot personality development

## Loading the Dataset

```{r load-data}
library(graniteR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)

data(emotion_full, package = "graniteR")
emotions <- emotion_full

glimpse(emotions)
```

The dataset contains:
- `text`: The text content
- `label`: Integer label (0-5)
- `label_name`: Emotion name (sadness, joy, love, anger, fear, surprise)

## Exploratory Data Analysis

### Class Distribution

```{r class-distribution, fig.height=6}
class_stats <- emotions |>
  count(label_name) |>
  mutate(
    percentage = n / sum(n) * 100,
    label_text = sprintf("%s\n%s (%.1f%%)", label_name, comma(n), percentage)
  ) |>
  arrange(desc(n))

ggplot(class_stats, aes(x = reorder(label_name, n), y = n, fill = label_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", comma(n), percentage)),
            hjust = -0.1, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma, limits = c(0, max(class_stats$n) * 1.15)) +
  coord_flip() +
  labs(
    title = "Emotion Distribution in Dataset",
    x = "",
    y = "Number of Samples"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold")
  )
```

The dataset shows class imbalance, with joy and sadness being most common, while surprise is relatively rare. This reflects real-world emotion distributions in text.

### Text Length Analysis

```{r length-analysis, fig.height=5}
emotions_analysis <- emotions |>
  mutate(
    word_count = stringr::str_count(text, "\\S+"),
    char_length = nchar(text)
  )

ggplot(emotions_analysis, aes(x = label_name, y = word_count, fill = label_name)) +
  geom_violin(alpha = 0.7, show.legend = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.5, outlier.alpha = 0.3, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma) +
  coord_cartesian(ylim = c(0, 50)) +
  labs(
    title = "Text Length Distribution by Emotion",
    x = "",
    y = "Word Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r length-stats}
emotions_analysis |>
  group_by(label_name) |>
  summarise(
    samples = n(),
    mean_words = round(mean(word_count), 1),
    median_words = round(median(word_count), 1)
  ) |>
  arrange(desc(samples)) |>
  knitr::kable(caption = "Text Statistics by Emotion")
```

### Sample Texts

```{r sample-texts}
cat("Sample texts by emotion:\n\n")
for (emotion in unique(emotions$label_name)) {
  cat(sprintf("** %s **\n", toupper(emotion)))
  emotions |>
    filter(label_name == emotion) |>
    slice_sample(n = 2) |>
    pull(text) |>
    cat(sep = "\n")
  cat("\n\n")
}
```

## Data Preparation

Split the data into training and testing sets:

```{r}
set.seed(42)

n <- nrow(emotions)
train_idx <- sample(n, size = floor(0.8 * n))

train_data <- emotions[train_idx, ]
test_data <- emotions[-train_idx, ]

cat(sprintf("Training samples: %s\n", comma(nrow(train_data))))
cat(sprintf("Testing samples: %s\n", comma(nrow(test_data))))
```

## Training a Classifier

Create and train a multi-class classifier for emotion detection:

```{r}
classifier <- classifier(num_labels = 6)

classifier <- classifier |>
  train(
    train_data,
    text,
    label,
    epochs = 5,
    batch_size = 8,
    learning_rate = 1e-3,
    validation_split = 0.2
  )
```

The classification head is trained while keeping the pretrained Granite model frozen. This approach works well for multi-class problems and prevents overfitting.

## Making Predictions

Apply the trained classifier to the test set:

```{r}
predictions <- predict(
  classifier,
  test_data,
  text,
  type = "class"
)

probabilities <- predict(
  classifier,
  test_data,
  text,
  type = "prob"
)

predictions |>
  select(text, label, prediction) |>
  mutate(
    true_emotion = c("sadness", "joy", "love", "anger", "fear", "surprise")[label + 1],
    pred_emotion = c("sadness", "joy", "love", "anger", "fear", "surprise")[prediction + 1]
  ) |>
  select(text, true_emotion, pred_emotion) |>
  head(10) |>
  knitr::kable()
```

## Model Evaluation

### Overall Performance

```{r metrics}
accuracy <- mean(predictions$prediction == predictions$label)

cat(sprintf("Overall Accuracy: %.2f%%\n", accuracy * 100))
```

### Confusion Matrix

```{r confusion-matrix, fig.height=7, fig.width=8}
emotion_names <- c("sadness", "joy", "love", "anger", "fear", "surprise")

conf_matrix_data <- predictions |>
  mutate(
    true_label = emotion_names[label + 1],
    pred_label = emotion_names[prediction + 1]
  ) |>
  count(true_label, pred_label) |>
  complete(true_label = emotion_names, pred_label = emotion_names, fill = list(n = 0))

ggplot(conf_matrix_data, aes(x = true_label, y = pred_label, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#f0f0f0", high = "#3498db", labels = comma) +
  labs(
    title = "Confusion Matrix: Emotion Detection",
    x = "True Emotion",
    y = "Predicted Emotion",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

### Per-Class Metrics

```{r per-class-metrics, fig.height=6}
calculate_metrics <- function(predictions, class_label) {
  tp <- sum(predictions$prediction == class_label & predictions$label == class_label)
  fp <- sum(predictions$prediction == class_label & predictions$label != class_label)
  fn <- sum(predictions$prediction != class_label & predictions$label == class_label)
  tn <- sum(predictions$prediction != class_label & predictions$label != class_label)

  precision <- if (tp + fp > 0) tp / (tp + fp) else 0
  recall <- if (tp + fn > 0) tp / (tp + fn) else 0
  f1 <- if (precision + recall > 0) 2 * precision * recall / (precision + recall) else 0

  data.frame(
    emotion = emotion_names[class_label + 1],
    precision = precision,
    recall = recall,
    f1 = f1
  )
}

metrics_df <- bind_rows(lapply(0:5, function(i) calculate_metrics(predictions, i)))

metrics_long <- metrics_df |>
  pivot_longer(cols = c(precision, recall, f1),
               names_to = "metric",
               values_to = "value")

ggplot(metrics_long, aes(x = emotion, y = value, fill = metric)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", value)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3) +
  scale_fill_manual(
    values = c("precision" = "#e74c3c", "recall" = "#3498db", "f1" = "#2ecc71"),
    labels = c("Precision", "Recall", "F1 Score")
  ) +
  scale_y_continuous(limits = c(0, 1.1), labels = percent) +
  labs(
    title = "Performance Metrics by Emotion",
    x = "",
    y = "Score",
    fill = "Metric"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r metrics-table}
metrics_df |>
  mutate(across(c(precision, recall, f1), ~sprintf("%.3f", .))) |>
  knitr::kable(caption = "Detailed Metrics by Emotion Class")
```

### Probability Distribution

```{r prob-distribution, fig.height=7}
prob_analysis <- probabilities |>
  mutate(
    true_emotion = emotion_names[test_data$label[1:n()] + 1],
    max_prob = pmax(prob_1, prob_2, prob_3, prob_4, prob_5, prob_6),
    predicted_class = max.col(cbind(prob_1, prob_2, prob_3, prob_4, prob_5, prob_6)) - 1,
    pred_emotion = emotion_names[predicted_class + 1],
    correct = predicted_class == test_data$label[1:n()]
  )

ggplot(prob_analysis, aes(x = max_prob, fill = correct)) +
  geom_histogram(bins = 30, alpha = 0.8) +
  facet_wrap(~true_emotion, ncol = 2) +
  scale_fill_manual(
    values = c("TRUE" = "#27ae60", "FALSE" = "#c0392b"),
    labels = c("Incorrect", "Correct")
  ) +
  labs(
    title = "Prediction Confidence Distribution by Emotion",
    x = "Maximum Probability (Confidence)",
    y = "Count",
    fill = "Prediction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )
```

## Error Analysis

### Misclassified Examples

```{r error-analysis}
misclassified <- predictions |>
  filter(prediction != label) |>
  mutate(
    true_emotion = emotion_names[label + 1],
    pred_emotion = emotion_names[prediction + 1]
  ) |>
  select(text, true_emotion, pred_emotion)

cat(sprintf("Misclassified samples: %s (%.2f%% of total)\n\n",
            comma(nrow(misclassified)),
            nrow(misclassified) / nrow(predictions) * 100))

cat("Examples of misclassified texts:\n\n")
misclassified |>
  slice_sample(n = min(5, nrow(misclassified))) |>
  knitr::kable(col.names = c("Text", "True", "Predicted"))
```

### Common Confusion Patterns

```{r confusion-patterns}
confusion_summary <- predictions |>
  filter(prediction != label) |>
  mutate(
    true_emotion = emotion_names[label + 1],
    pred_emotion = emotion_names[prediction + 1],
    confusion = paste(true_emotion, "→", pred_emotion)
  ) |>
  count(confusion, sort = TRUE) |>
  head(10)

cat("Most common misclassifications:\n\n")
confusion_summary |>
  knitr::kable(col.names = c("Confusion Pattern", "Count"))
```

## Deployment Considerations

For production deployment of emotion detection models:

1. **Class Imbalance**: Consider using weighted loss functions or oversampling rare emotions
2. **Threshold Tuning**: Adjust confidence thresholds based on application requirements
3. **Context Awareness**: Emotion can be context-dependent; consider multi-sentence context
4. **Multilingual Support**: Train separate models for different languages or use multilingual base models
5. **Regular Updates**: Retrain as language patterns and emoji usage evolve

### Example: High-Confidence Filtering

```{r confidence-filtering}
high_confidence <- prob_analysis |>
  filter(max_prob > 0.8) |>
  select(text, true_emotion, pred_emotion, max_prob, correct)

cat(sprintf("High-confidence predictions (>80%%): %s (%.1f%% of total)\n",
            comma(nrow(high_confidence)),
            nrow(high_confidence) / nrow(prob_analysis) * 100))

cat(sprintf("Accuracy on high-confidence predictions: %.2f%%\n",
            mean(high_confidence$correct) * 100))
```

## Conclusion

This vignette demonstrated using graniteR for multi-class emotion detection, achieving strong classification performance across six emotion categories. The classification head training approach provides excellent results while being efficient and preventing overfitting.

Key takeaways:
- Multi-class classification requires careful handling of class imbalance
- Confusion matrices reveal which emotions are most difficult to distinguish
- Confidence-based filtering can improve precision at the cost of coverage
- The frozen-encoder approach works well even with imbalanced multi-class data

## References

- Granite Embedding R2 Models (2025): https://arxiv.org/html/2508.21085v1
- Emotion Dataset: https://huggingface.co/datasets/dair-ai/emotion
- IBM Granite Models: https://huggingface.co/ibm-granite
- Transformers Library: https://huggingface.co/docs/transformers
