---
title: "Emotion Detection with graniteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Emotion Detection with graniteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 10,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)

# Check if pre-trained models are available by directly checking file existence
models_available <- tryCatch({
  # Check in inst/extdata/models (for installed package)
  pkg_models_dir <- system.file("extdata", "models", package = "graniteR")
  if (nzchar(pkg_models_dir)) {
    file.exists(file.path(pkg_models_dir, "emotion_standard_config.rds")) &&
    file.exists(file.path(pkg_models_dir, "emotion_moe_config.rds"))
  } else {
    FALSE
  }
}, error = function(e) FALSE)
```

## Introduction

This vignette demonstrates using graniteR for multi-class emotion detection. We use the emotion dataset from HuggingFace (dair-ai/emotion), which contains text samples labeled with six basic emotions.

The dataset includes 20,000 labeled examples for training robust emotion classifiers. We demonstrate two approaches:

1. **Standard Classifier**: Frozen pretrained model with trainable classification head
2. **Mixture of Experts (MoE)**: Multiple expert networks with specialized routing

## Loading the Dataset

```{r load-data}
library(graniteR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)

data(emotion_full, package = "graniteR")
emotions <- emotion_full

glimpse(emotions)
```

The dataset contains:
- `text`: The text content
- `label`: Integer label (0-5)
- `label_name`: Emotion name (sadness, joy, love, anger, fear, surprise)

## Class Distribution

```{r class-distribution, fig.height=5}
class_stats <- emotions |>
  count(label_name) |>
  mutate(
    percentage = n / sum(n) * 100,
    label_text = sprintf("%s\n%s (%.1f%%)", label_name, comma(n), percentage)
  ) |>
  arrange(desc(n))

 ggplot(class_stats, aes(x = reorder(label_name, n), y = n, fill = label_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", comma(n), percentage)),
            hjust = -0.1, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma, limits = c(0, max(class_stats$n) * 1.15)) +
  coord_flip() +
  labs(
    title = "Emotion Distribution in Dataset",
    x = "",
    y = "Number of Samples"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold")
  )
```

The dataset shows class imbalance, with joy and sadness being most common.

## Data Preparation

```{r data-prep}
set.seed(42)

n <- nrow(emotions)
train_idx <- sample(n, size = floor(0.8 * n))

train_data <- emotions[train_idx, ]
test_data <- emotions[-train_idx, ]

cat(sprintf("Training samples: %s\n", comma(nrow(train_data))))
cat(sprintf("Testing samples: %s\n", comma(nrow(test_data))))
```

# Approach 1: Standard Classifier

## Training

```{r training-standard, eval=FALSE}
clf_standard <- classifier(num_labels = 6)

clf_standard <- clf_standard |>
  train(
    train_data,
    text,
    label,
    epochs = 5,
    batch_size = 50,
    learning_rate = 1e-3,
    validation_split = 0.2
  )

save_classifier(clf_standard, file = "models/emotion_standard")
```

The classification head is trained while keeping the pretrained Granite model frozen. This approach prevents overfitting and requires less training time.

## Loading Pre-trained Model

For this vignette, we load a pre-trained model. The `find_model()` function automatically locates the model in the package or your local download directory:

```{r load-model-standard, eval=models_available}
clf_standard <- load_classifier(find_model("emotion_standard"))
```

## Making Predictions

```{r predictions-standard, eval=models_available}
pred_standard <- predict(
  clf_standard,
  test_data,
  text,
  type = "class"
)

pred_standard |>
  select(text, label, prediction) |>
  mutate(
    true_emotion = c("sadness", "joy", "love", "anger", "fear", "surprise")[label + 1],
    pred_emotion = c("sadness", "joy", "love", "anger", "fear", "surprise")[prediction + 1]
  ) |>
  select(text, true_emotion, pred_emotion) |>
  head(10) |>
  knitr::kable()
```

## Model Evaluation

```{r metrics-standard, eval=models_available}
accuracy_standard <- mean(pred_standard$prediction == pred_standard$label)
cat(sprintf("Standard Classifier Accuracy: %.2f%%\n", accuracy_standard * 100))
```

### Confusion Matrix

```{r confusion-matrix-standard, fig.height=6, fig.width=7, eval=models_available}
emotion_names <- c("sadness", "joy", "love", "anger", "fear", "surprise")

conf_matrix_std <- pred_standard |>
  mutate(
    true_label = emotion_names[label + 1],
    pred_label = emotion_names[prediction + 1]
  ) |>
  count(true_label, pred_label) |>
  complete(true_label = emotion_names, pred_label = emotion_names, fill = list(n = 0))

 ggplot(conf_matrix_std, aes(x = true_label, y = pred_label, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#f0f0f0", high = "#3498db", labels = comma) +
  labs(
    title = "Confusion Matrix: Standard Classifier",
    x = "True Emotion",
    y = "Predicted Emotion",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

# Approach 2: Mixture of Experts (MoE)

The MoE approach uses multiple specialized expert networks with a gating mechanism that routes examples to the most appropriate expert.

## Training

```{r training-moe, eval=FALSE}
clf_moe <- moe_classifier(
  num_labels = 6,
  num_experts = 6,
  freeze_backbone = FALSE,
  dropout = 0.2,
  expert_depth = 2
)

clf_moe <- train_moe(
  clf_moe,
  train_data,
  text,
  label,
  epochs = 5,
  batch_size = 8,
  learning_rate = 2e-5,
  validation_split = 0.2
)

save_classifier(clf_moe, file = "models/emotion_moe")
```

## Loading Pre-trained MoE Model

```{r load-model-moe, eval=models_available}
clf_moe <- load_classifier(find_model("emotion_moe"))
```

## Making Predictions

```{r predictions-moe, eval=models_available}
pred_moe <- predict(
  clf_moe,
  test_data,
  text,
  type = "class"
)

pred_moe |>
  select(text, label, prediction) |>
  mutate(
    true_emotion = emotion_names[label + 1],
    pred_emotion = emotion_names[prediction + 1]
  ) |>
  select(text, true_emotion, pred_emotion) |>
  head(10) |>
  knitr::kable()
```

## Model Evaluation

```{r metrics-moe, eval=models_available}
accuracy_moe <- mean(pred_moe$prediction == pred_moe$label)
cat(sprintf("MoE Classifier Accuracy: %.2f%%\n", accuracy_moe * 100))
```

### Confusion Matrix

```{r confusion-matrix-moe, fig.height=6, fig.width=7, eval=models_available}
conf_matrix_moe <- pred_moe |>
  mutate(
    true_label = emotion_names[label + 1],
    pred_label = emotion_names[prediction + 1]
  ) |>
  count(true_label, pred_label) |>
  complete(true_label = emotion_names, pred_label = emotion_names, fill = list(n = 0))

 ggplot(conf_matrix_moe, aes(x = true_label, y = pred_label, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#f0f0f0", high = "#9b59b6", labels = comma) +
  labs(
    title = "Confusion Matrix: MoE Classifier",
    x = "True Emotion",
    y = "Predicted Emotion",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

# Comparing Approaches

```{r comparison, eval=models_available}
comparison <- data.frame(
  Approach = c("Standard Classifier", "MoE Classifier"),
  Accuracy = c(accuracy_standard, accuracy_moe),
  Training_Time = c("Fast", "Moderate"),
  Model_Size = c("Small", "Large"),
  Best_For = c("General use, efficiency", "Complex patterns, high accuracy")
)

knitr::kable(comparison, digits = 4, caption = "Comparison of Classification Approaches")
```

### Per-Class Metrics Comparison

```{r per-class-comparison, fig.height=7}
calculate_metrics <- function(predictions, class_label, approach_name) {
  tp <- sum(predictions$prediction == class_label & predictions$label == class_label)
  fp <- sum(predictions$prediction == class_label & predictions$label != class_label)
  fn <- sum(predictions$prediction != class_label & predictions$label == class_label)

  precision <- if (tp + fp > 0) tp / (tp + fp) else 0
  recall <- if (tp + fn > 0) tp / (tp + fn) else 0
  f1 <- if (precision + recall > 0) 2 * precision * recall / (precision + recall) else 0

  data.frame(
    approach = approach_name,
    emotion = emotion_names[class_label + 1],
    precision = precision,
    recall = recall,
    f1 = f1
  )
}

metrics_standard <- bind_rows(lapply(0:5, function(i)
  calculate_metrics(pred_standard, i, "Standard")))

metrics_moe <- bind_rows(lapply(0:5, function(i)
  calculate_metrics(pred_moe, i, "MoE")))

metrics_combined <- bind_rows(metrics_standard, metrics_moe) |>
  pivot_longer(cols = c(precision, recall, f1),
               names_to = "metric",
               values_to = "value")

ggplot(metrics_combined, aes(x = emotion, y = value, fill = approach)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~ metric, ncol = 1) +
  scale_fill_manual(
    values = c("Standard" = "#3498db", "MoE" = "#9b59b6"),
    labels = c("Standard Classifier", "MoE Classifier")
  ) +
  scale_y_continuous(limits = c(0, 1), labels = percent) +
  labs(
    title = "Performance Comparison: Standard vs MoE",
    x = "",
    y = "Score",
    fill = "Approach"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 12, face = "bold")
  )
```

## Conclusion

This vignette demonstrated two approaches for multi-class emotion detection with graniteR:

**Standard Classifier:**
- Fast training (frozen backbone)
- Efficient inference
- Good baseline performance
- Ideal for most use cases

**MoE Classifier:**
- Longer training time (fine-tuned backbone)
- Multiple specialized experts
- Potentially higher accuracy
- Better for complex, nuanced classification tasks

Both approaches leverage IBM's Granite embedding models and achieve strong performance on the emotion detection task.

## References

- Granite Embedding R2 Models (2025): https://arxiv.org/html/2508.21085v1
- Emotion Dataset: https://huggingface.co/datasets/dair-ai/emotion
- IBM Granite Models: https://huggingface.co/ibm-granite
