---
title: "Emotion Detection with graniteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Emotion Detection with graniteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 10,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates using graniteR for multi-class emotion detection. We use the emotion dataset from HuggingFace (dair-ai/emotion), which contains text samples labeled with six basic emotions.

The dataset includes 20,000 labeled examples for training robust emotion classifiers. Key features:

- Multi-class text classification (6 emotions)
- Transfer learning with frozen pretrained models
- Fast training (classification head only)

## Loading the Dataset

```{r load-data}
library(graniteR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)

data(emotion_full, package = "graniteR")
emotions <- emotion_full

glimpse(emotions)
```

The dataset contains:
- `text`: The text content
- `label`: Integer label (0-5)
- `label_name`: Emotion name (sadness, joy, love, anger, fear, surprise)

## Class Distribution

```{r class-distribution, fig.height=5}
class_stats <- emotions |>
  count(label_name) |>
  mutate(
    percentage = n / sum(n) * 100,
    label_text = sprintf("%s\n%s (%.1f%%)", label_name, comma(n), percentage)
  ) |>
  arrange(desc(n))

 ggplot(class_stats, aes(x = reorder(label_name, n), y = n, fill = label_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", comma(n), percentage)),
            hjust = -0.1, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma, limits = c(0, max(class_stats$n) * 1.15)) +
  coord_flip() +
  labs(
    title = "Emotion Distribution in Dataset",
    x = "",
    y = "Number of Samples"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold")
  )
```

The dataset shows class imbalance, with joy and sadness being most common.

## Data Preparation

```{r data-prep}
set.seed(42)

n <- nrow(emotions)
train_idx <- sample(n, size = floor(0.8 * n))

train_data <- emotions[train_idx, ]
test_data <- emotions[-train_idx, ]

cat(sprintf("Training samples: %s\n", comma(nrow(train_data))))
cat(sprintf("Testing samples: %s\n", comma(nrow(test_data))))
```

## Training a Classifier

```{r training, eval=FALSE}
classifier <- classifier(num_labels = 6)

classifier <- classifier |>
  train(
    train_data,
    text,
    label,
    epochs = 5,
    batch_size = 50,
    learning_rate = 1e-3,
    validation_split = 0.2
  )

save_classifier(classifier, file = "models/emotion_standard")
```

The classification head is trained while keeping the pretrained Granite model frozen. This approach prevents overfitting and requires less training time.

## Loading Pre-trained Model

For this vignette, we load a pre-trained model to show predictions:

```{r load-model, eval=TRUE}
classifier <- load_classifier("models/emotion_standard")
```

## Making Predictions

```{r predictions}
predictions <- predict(
  classifier,
  test_data,
  text,
  type = "class"
)

predictions |>
  select(text, label, prediction) |>
  mutate(
    true_emotion = c("sadness", "joy", "love", "anger", "fear", "surprise")[label + 1],
    pred_emotion = c("sadness", "joy", "love", "anger", "fear", "surprise")[prediction + 1]
  ) |>
  select(text, true_emotion, pred_emotion) |>
  head(10) |>
  knitr::kable()
```

## Model Evaluation

```{r metrics}
accuracy <- mean(predictions$prediction == predictions$label)
cat(sprintf("Overall Accuracy: %.2f%%\n", accuracy * 100))
```

### Confusion Matrix

```{r confusion-matrix, fig.height=6, fig.width=7}
emotion_names <- c("sadness", "joy", "love", "anger", "fear", "surprise")

conf_matrix_data <- predictions |>
  mutate(
    true_label = emotion_names[label + 1],
    pred_label = emotion_names[prediction + 1]
  ) |>
  count(true_label, pred_label) |>
  complete(true_label = emotion_names, pred_label = emotion_names, fill = list(n = 0))

ggplot(conf_matrix_data, aes(x = true_label, y = pred_label, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#f0f0f0", high = "#3498db", labels = comma) +
  labs(
    title = "Confusion Matrix: Emotion Detection",
    x = "True Emotion",
    y = "Predicted Emotion",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

### Per-Class Metrics

```{r per-class-metrics, fig.height=6}
calculate_metrics <- function(predictions, class_label) {
  tp <- sum(predictions$prediction == class_label & predictions$label == class_label)
  fp <- sum(predictions$prediction == class_label & predictions$label != class_label)
  fn <- sum(predictions$prediction != class_label & predictions$label == class_label)

  precision <- if (tp + fp > 0) tp / (tp + fp) else 0
  recall <- if (tp + fn > 0) tp / (tp + fn) else 0
  f1 <- if (precision + recall > 0) 2 * precision * recall / (precision + recall) else 0

  data.frame(
    emotion = emotion_names[class_label + 1],
    precision = precision,
    recall = recall,
    f1 = f1
  )
}

metrics_df <- bind_rows(lapply(0:5, function(i) calculate_metrics(predictions, i)))

metrics_long <- metrics_df |>
  pivot_longer(cols = c(precision, recall, f1),
               names_to = "metric",
               values_to = "value")

ggplot(metrics_long, aes(x = emotion, y = value, fill = metric)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", value)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3) +
  scale_fill_manual(
    values = c("precision" = "#e74c3c", "recall" = "#3498db", "f1" = "#2ecc71"),
    labels = c("Precision", "Recall", "F1 Score")
  ) +
  scale_y_continuous(limits = c(0, 1.1), labels = percent) +
  labs(
    title = "Performance Metrics by Emotion",
    x = "",
    y = "Score",
    fill = "Metric"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r metrics-table}
metrics_df |>
  mutate(across(c(precision, recall, f1), ~sprintf("%.3f", .))) |>
  knitr::kable(caption = "Detailed Metrics by Emotion Class")
```

## Conclusion

This vignette demonstrated using graniteR for multi-class emotion detection, achieving strong classification performance across six emotion categories. The frozen-encoder approach provides excellent results while being efficient.

Key takeaways:
- Multi-class classification with graniteR is straightforward
- Frozen pretrained models work well for text classification
- Training is fast (classification head only)
- Results are interpretable through confusion matrices and per-class metrics

## References

- Granite Embedding R2 Models (2025): https://arxiv.org/html/2508.21085v1
- Emotion Dataset: https://huggingface.co/datasets/dair-ai/emotion
- IBM Granite Models: https://huggingface.co/ibm-granite
