---
title: "Mixture of Experts Classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixture of Experts Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 10,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates the Mixture of Experts (MoE) architecture for text classification in graniteR. MoE uses multiple specialized expert networks whose outputs are dynamically weighted based on the input.

**Important**: MoE requires more training data and compute than standard classifiers. Always start with the standard classifier first.

## Understanding Mixture of Experts

### Architecture

A standard classifier:
```
Input → Backbone → Single Head → Output
```

A Mixture of Experts:
```
Input → Backbone → ┌ Expert 1 ┐
                   ├ Expert 2 ├ → Weighted Sum → Output
                   ├ Expert 3 ├
                   └ Expert 4 ┘
                        ↑
                   Gating Network
```

### When to Use MoE

MoE may help when:
- Large training datasets (20K+ samples)
- Complex multi-class problems (4+ classes)
- Standard classifier has plateaued
- Willing to fine-tune entire backbone

Use standard classifier when:
- Small datasets (<10K samples)
- Binary classification
- Limited compute budget
- Quick experimentation needed

## Loading the Dataset

```{r load-data}
library(graniteR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)

data(emotion_full, package = "graniteR")
emotions <- emotion_full

set.seed(42)
n <- nrow(emotions)
train_idx <- sample(n, size = floor(0.8 * n))

train_data <- emotions[train_idx, ]
test_data <- emotions[-train_idx, ]

cat(sprintf("Training samples: %s\n", comma(nrow(train_data))))
cat(sprintf("Testing samples: %s\n", comma(nrow(test_data))))
```

## Training Standard vs MoE Classifiers

### Standard Classifier

```{r standard-training, eval=FALSE}
clf_standard <- classifier(num_labels = 6)

clf_standard <- clf_standard |>
  train(
    train_data,
    text,
    label,
    epochs = 5,
    batch_size = 8,
    learning_rate = 1e-3,
    validation_split = 0.2
  )

save_classifier(clf_standard, file = "models/emotion_standard")
```

### MoE Classifier

```{r moe-training, eval=FALSE}
clf_moe <- moe_classifier(
  num_labels = 6,
  num_experts = 6,
  freeze_backbone = FALSE,
  dropout = 0.2,
  expert_depth = 2
)

clf_moe <- clf_moe |>
  train_moe(
    train_data,
    text,
    label,
    epochs = 5,
    batch_size = 8,
    learning_rate = 2e-5,
    validation_split = 0.2
  )

save_classifier(clf_moe, file = "models/emotion_moe")
```

**Key Differences**:
- `freeze_backbone = FALSE`: Full fine-tuning required for MoE
- Lower learning rate (2e-5 vs 1e-3): Required when fine-tuning backbone
- `expert_depth = 2`: Deeper expert networks

MoE shows additional loss components:
- **cls**: Classification loss
- **lb**: Load balancing loss (encourages expert diversity)

## Loading Pre-trained Models

```{r load-models, eval=TRUE}
clf_standard <- load_classifier("models/emotion_standard")
clf_moe <- load_classifier("models/emotion_moe")
```

## Making Predictions

```{r predictions}
pred_standard <- predict(clf_standard, test_data, text, type = "class")
pred_moe <- predict(clf_moe, test_data, text, type = "class")

expert_weights <- predict(clf_moe, test_data, text, type = "expert_weights")
```

The `expert_weights` output shows which experts were most active for each input.

## Performance Comparison

```{r performance-comparison}
emotion_names <- c("sadness", "joy", "love", "anger", "fear", "surprise")

calculate_metrics <- function(predictions) {
  accuracy <- mean(predictions$prediction == predictions$label)

  per_class <- lapply(0:5, function(class_id) {
    tp <- sum(predictions$prediction == class_id & predictions$label == class_id)
    fp <- sum(predictions$prediction == class_id & predictions$label != class_id)
    fn <- sum(predictions$prediction != class_id & predictions$label == class_id)

    precision <- if (tp + fp > 0) tp / (tp + fp) else 0
    recall <- if (tp + fn > 0) tp / (tp + fn) else 0
    f1 <- if (precision + recall > 0) 2 * precision * recall / (precision + recall) else 0

    data.frame(
      emotion = emotion_names[class_id + 1],
      precision = precision,
      recall = recall,
      f1 = f1
    )
  }) |> bind_rows()

  list(
    accuracy = accuracy,
    per_class = per_class
  )
}

metrics_standard <- calculate_metrics(pred_standard)
metrics_moe <- calculate_metrics(pred_moe)

comparison <- data.frame(
  Model = c("Standard", "MoE"),
  Accuracy = c(metrics_standard$accuracy, metrics_moe$accuracy),
  `Avg F1` = c(
    mean(metrics_standard$per_class$f1),
    mean(metrics_moe$per_class$f1)
  )
)

cat("Overall Performance Comparison:\n\n")
comparison |>
  mutate(across(where(is.numeric), ~sprintf("%.4f", .))) |>
  knitr::kable()
```

```{r per-class-comparison, fig.height=6}
per_class_comparison <- bind_rows(
  metrics_standard$per_class |> mutate(Model = "Standard"),
  metrics_moe$per_class |> mutate(Model = "MoE")
) |>
  pivot_longer(cols = c(precision, recall, f1),
               names_to = "metric",
               values_to = "value")

ggplot(per_class_comparison, aes(x = emotion, y = value, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~metric, ncol = 1) +
  scale_fill_manual(values = c("Standard" = "#3498db", "MoE" = "#e74c3c")) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Per-Class Performance: Standard vs MoE",
    x = "",
    y = "Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )
```

## Analyzing Expert Specialization

```{r expert-specialization, fig.height=6}
expert_analysis <- expert_weights |>
  mutate(
    true_emotion = emotion_names[test_data$label[1:n()] + 1],
    primary_expert = max.col(select(expert_weights, starts_with("expert_"))) - 1
  )

expert_class_matrix <- expert_analysis |>
  count(true_emotion, primary_expert) |>
  complete(true_emotion = emotion_names, primary_expert = 0:5, fill = list(n = 0)) |>
  mutate(primary_expert = paste0("Expert ", primary_expert))

ggplot(expert_class_matrix, aes(x = true_emotion, y = primary_expert, fill = n)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = n), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "#f0f0f0", high = "#3498db", labels = comma) +
  labs(
    title = "Expert Specialization by Emotion Class",
    x = "True Emotion",
    y = "",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

## Best Practices

1. **Always Start Simple**: Train standard classifier first
2. **Set Realistic Expectations**: MoE may only improve by 1-3%
3. **Full Fine-Tuning Required**: MoE needs unfrozen backbone
4. **More Data Needed**: Aim for 5-10K samples per class
5. **Lower Learning Rates**: Use 2e-5 to 5e-5
6. **Monitor Load Balancing**: Check lb loss during training
7. **Expert Count**: Start with 3-4 experts

## Conclusion

Mixture of Experts is an advanced technique with realistic tradeoffs:

**When MoE Helps:**
- Large datasets (50K+ samples)
- Complex multi-class problems (6+ classes)
- Standard classifier has plateaued

**When to Use Standard Classifier:**
- Small datasets (<20K samples)
- Binary classification
- Limited compute budget

**Realistic Expectations:**
- MoE may improve accuracy by 1-3%
- Requires 2-3x more training time
- Standard classifier often competitive

**Bottom Line:** Use MoE as an advanced optimization step after exhausting simpler approaches.

## References

- Mixture of Experts: https://arxiv.org/abs/1701.06538
- Granite Embedding R2 Models: https://arxiv.org/html/2508.21085v1
- IBM Granite Models: https://huggingface.co/ibm-granite
