---
title: "Sentiment Analysis with graniteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sentiment Analysis with graniteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 10,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates using graniteR for binary sentiment analysis on movie reviews using the IMDB dataset. We demonstrate two approaches:

1. **Standard Classifier**: Frozen pretrained model with trainable classification head
2. **Mixture of Experts (MoE)**: Multiple expert networks with specialized routing

The IMDB dataset includes 50,000 movie reviews labeled as positive or negative, making it an ideal benchmark for sentiment classification.

## Loading the Dataset

```{r load-data}
library(graniteR)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(scales)

data(sentiment_imdb_full, package = "graniteR")
reviews <- sentiment_imdb_full

glimpse(reviews)
```

The dataset contains:
- `text`: The movie review content
- `label`: Binary label (0 = negative, 1 = positive)
- `label_name`: Sentiment name (negative or positive)

## Class Distribution

```{r class-distribution, fig.height=5}
class_stats <- reviews |>
  count(label_name) |>
  mutate(
    percentage = n / sum(n) * 100,
    label_text = sprintf("%s\n%s (%.1f%%)", label_name, comma(n), percentage)
  )

ggplot(class_stats, aes(x = "", y = n, fill = label_name)) +
  geom_col(width = 1, color = "white", size = 1.5) +
  geom_text(aes(label = label_text),
            position = position_stack(vjust = 0.5),
            size = 6, fontface = "bold", color = "white") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("negative" = "#e74c3c", "positive" = "#2ecc71")) +
  theme_void() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  ) +
  labs(title = "Dataset Class Distribution")
```

The IMDB dataset is perfectly balanced with equal numbers of positive and negative reviews.

## Data Preparation

```{r data-prep}
set.seed(42)

n <- nrow(reviews)
train_idx <- sample(n, size = floor(0.8 * n))

train_data <- reviews[train_idx, ]
test_data <- reviews[-train_idx, ]

cat(sprintf("Training samples: %s\n", comma(nrow(train_data))))
cat(sprintf("Testing samples: %s\n", comma(nrow(test_data))))
```

# Approach 1: Standard Classifier

## Training

```{r training-standard, eval=FALSE}
clf_standard <- classifier(num_labels = 2)

clf_standard <- clf_standard |>
  train(
    train_data,
    text,
    label,
    epochs = 3,
    batch_size = 8,
    learning_rate = 1e-3,
    validation_split = 0.2
  )

save_classifier(clf_standard, file = "models/sentiment_standard")
```

The classification head is trained while keeping the pretrained Granite model frozen.

## Loading Pre-trained Model

```{r load-model-standard, eval=FALSE}
clf_standard <- load_classifier(find_model("sentiment_standard"))
```

## Making Predictions

```{r predictions-standard}
pred_standard <- predict(
  clf_standard,
  test_data,
  text,
  type = "class"
)

pred_standard |>
  mutate(
    review_preview = substr(text, 1, 80),
    true_sentiment = c("negative", "positive")[label + 1],
    pred_sentiment = c("negative", "positive")[prediction + 1]
  ) |>
  select(review_preview, true_sentiment, pred_sentiment) |>
  head(10) |>
  knitr::kable()
```

## Model Evaluation

```{r metrics-standard}
tp_std <- sum(pred_standard$prediction == 1 & pred_standard$label == 1)
fp_std <- sum(pred_standard$prediction == 1 & pred_standard$label == 0)
fn_std <- sum(pred_standard$prediction == 0 & pred_standard$label == 1)
tn_std <- sum(pred_standard$prediction == 0 & pred_standard$label == 0)

accuracy_standard <- (tp_std + tn_std) / (tp_std + tn_std + fp_std + fn_std)
precision_standard <- tp_std / (tp_std + fp_std)
recall_standard <- tp_std / (tp_std + fn_std)
f1_standard <- 2 * (precision_standard * recall_standard) / (precision_standard + recall_standard)

cat(sprintf("Standard Classifier Accuracy: %.2f%%\n", accuracy_standard * 100))
```

### Confusion Matrix

```{r confusion-matrix-standard, fig.height=6}
conf_matrix_std <- data.frame(
  Predicted = c("Negative", "Negative", "Positive", "Positive"),
  Actual = c("Negative", "Positive", "Negative", "Positive"),
  Count = c(tn_std, fn_std, fp_std, tp_std)
) |>
  mutate(
    Percentage = Count / sum(Count) * 100,
    Label = sprintf("%s\n(%.1f%%)", comma(Count), Percentage)
  )

ggplot(conf_matrix_std, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(color = "white", size = 2) +
  geom_text(aes(label = Label), size = 7, fontface = "bold") +
  scale_fill_gradient(low = "#ecf0f1", high = "#3498db", labels = comma) +
  labs(
    title = "Confusion Matrix: Standard Classifier",
    subtitle = sprintf("Accuracy: %.2f%%", accuracy_standard * 100),
    x = "Actual Sentiment",
    y = "Predicted Sentiment"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text = element_text(size = 12),
    legend.position = "right"
  ) +
  coord_fixed()
```

# Approach 2: Mixture of Experts (MoE)

The MoE approach uses multiple specialized expert networks with a gating mechanism that routes examples to the most appropriate expert.

## Training

```{r training-moe, eval=FALSE}
clf_moe <- moe_classifier(
  num_labels = 2,
  num_experts = 3,
  freeze_backbone = FALSE,
  dropout = 0.2,
  expert_depth = 2
)

clf_moe <- train_moe(
  clf_moe,
  train_data,
  text,
  label,
  epochs = 3,
  batch_size = 8
)

save_classifier(clf_moe, file = "models/sentiment_moe")
```

## Loading Pre-trained MoE Model

```{r load-model-moe, eval=FALSE}
clf_moe <- load_classifier(find_model("sentiment_moe"))
```

## Making Predictions

```{r predictions-moe}
pred_moe <- predict(
  clf_moe,
  test_data,
  text,
  type = "class"
)

pred_moe |>
  mutate(
    review_preview = substr(text, 1, 80),
    true_sentiment = c("negative", "positive")[label + 1],
    pred_sentiment = c("negative", "positive")[prediction + 1]
  ) |>
  select(review_preview, true_sentiment, pred_sentiment) |>
  head(10) |>
  knitr::kable()
```

## Model Evaluation

```{r metrics-moe}
tp_moe <- sum(pred_moe$prediction == 1 & pred_moe$label == 1)
fp_moe <- sum(pred_moe$prediction == 1 & pred_moe$label == 0)
fn_moe <- sum(pred_moe$prediction == 0 & pred_moe$label == 1)
tn_moe <- sum(pred_moe$prediction == 0 & pred_moe$label == 0)

accuracy_moe <- (tp_moe + tn_moe) / (tp_moe + tn_moe + fp_moe + fn_moe)
precision_moe <- tp_moe / (tp_moe + fp_moe)
recall_moe <- tp_moe / (tp_moe + fn_moe)
f1_moe <- 2 * (precision_moe * recall_moe) / (precision_moe + recall_moe)

cat(sprintf("MoE Classifier Accuracy: %.2f%%\n", accuracy_moe * 100))
```

### Confusion Matrix

```{r confusion-matrix-moe, fig.height=6}
conf_matrix_moe <- data.frame(
  Predicted = c("Negative", "Negative", "Positive", "Positive"),
  Actual = c("Negative", "Positive", "Negative", "Positive"),
  Count = c(tn_moe, fn_moe, fp_moe, tp_moe)
) |>
  mutate(
    Percentage = Count / sum(Count) * 100,
    Label = sprintf("%s\n(%.1f%%)", comma(Count), Percentage)
  )

ggplot(conf_matrix_moe, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(color = "white", size = 2) +
  geom_text(aes(label = Label), size = 7, fontface = "bold") +
  scale_fill_gradient(low = "#ecf0f1", high = "#9b59b6", labels = comma) +
  labs(
    title = "Confusion Matrix: MoE Classifier",
    subtitle = sprintf("Accuracy: %.2f%%", accuracy_moe * 100),
    x = "Actual Sentiment",
    y = "Predicted Sentiment"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text = element_text(size = 12),
    legend.position = "right"
  ) +
  coord_fixed()
```

# Comparing Approaches

```{r comparison}
comparison <- data.frame(
  Approach = c("Standard Classifier", "MoE Classifier"),
  Accuracy = c(accuracy_standard, accuracy_moe),
  Precision = c(precision_standard, precision_moe),
  Recall = c(recall_standard, recall_moe),
  F1_Score = c(f1_standard, f1_moe),
  Training_Time = c("Fast", "Moderate"),
  Model_Size = c("Small", "Large")
)

knitr::kable(comparison, digits = 4, caption = "Comparison of Classification Approaches")
```

### Performance Comparison

```{r performance-comparison, fig.height=6}
metrics_comparison <- data.frame(
  Approach = rep(c("Standard", "MoE"), each = 4),
  Metric = rep(c("Accuracy", "Precision", "Recall", "F1 Score"), 2),
  Value = c(accuracy_standard, precision_standard, recall_standard, f1_standard,
            accuracy_moe, precision_moe, recall_moe, f1_moe)
)

ggplot(metrics_comparison, aes(x = Metric, y = Value, fill = Approach)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", Value)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  scale_fill_manual(
    values = c("Standard" = "#3498db", "MoE" = "#9b59b6"),
    labels = c("Standard Classifier", "MoE Classifier")
  ) +
  scale_y_continuous(limits = c(0, 1.1), labels = percent) +
  labs(
    title = "Performance Comparison: Standard vs MoE",
    x = "",
    y = "Score",
    fill = "Approach"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "top"
  )
```

## Conclusion

This vignette demonstrated two approaches for binary sentiment analysis on movie reviews with graniteR:

**Standard Classifier:**
- Fast training (frozen backbone)
- Efficient inference
- Good baseline performance
- Ideal for most use cases

**MoE Classifier:**
- Longer training time (fine-tuned backbone)
- Multiple specialized experts
- Potentially higher accuracy
- Better for complex, nuanced sentiment

Both approaches leverage IBM's Granite embedding models and achieve strong performance on the IMDB benchmark dataset.

## References

- Granite Embedding R2 Models (2025): https://arxiv.org/html/2508.21085v1
- IMDB Dataset: https://huggingface.co/datasets/imdb
- IBM Granite Models: https://huggingface.co/ibm-granite
